A general and efficient representation of Ancestral Recombination Graphs

### 2022-06-06 paper.tex
Fixing requirements.txt for pip3

% NOTE: it might be helpful to have a drawing of the ARG process
% in-progress here to help us talk about it. What is the full state
% that is remembered as we go through the process? It does help to think
% about the process as choosing edges to merge, not nodes.
% We could also refer to Fig.~\ref{fig-arg-data-structure}B if we want a simple picture
% to go with this
In the Griffiths formulation, each edge in the graph corresponds to an extant
lineage and nodes are events in the process (the initial $n$ leaf nodes are
``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events happen at a rate of
$k \rho$. Here, we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMRCA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. Later, we will show how the graph structure and breakpoints
associated with
recombination nodes provides sufficient information to later recover the marginal
trees and calculate sampling probabilities. However, this process traces out
a larger graph than the Hudson algorithm (see the ``\nameref{Ancestry_resolution}'' section),
and is consequently considerably less efficient \citep{wiuf1999ancestry}.

@@

The idea of using branching-coalescing stochastic processes
to model genetic ancestral trees has also found application in natural selection.
The Ancestral Selection Graph (ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses dynamics identical to the ``big'' ARG to simulate an ensemble of correlated
potential ancestral trees. Weak genic selection is incorporated by sampling a
true ancestry from the ensemble in a non-uniform way.
There is no ASG analogue of the more computationally tractable ``little'' ARG,
though some gains in tractability can be made by considering typed
lineages~\citep{etheridge2009coalescent} or by leveraging perfect simulation
techniques when recurrent mutation is present\citep{fearnhead2001perfect}.
Extensions of the ASG have been developed to frequency-dependent
selection~\citep{neuhauser1999ancestral, gonzalezcasanova2018duality},
unlinked chromosomes~\citep{fearnhead2003ancestral}, recombining
loci whereupon branching is due to both selection and
recombination~\citep{donnelly1999genealogical}, and high fecundity
reproduction~\citep{gonzalezcasanova2018duality, koskela2019robust}.

@@

The term \emph{phylogenetic network} has been broadly accepted to describe the most general
type of genealogical graph, defined by \citet{huson2010phylogenetic} as ``any graph used to
represent evolutionary relationships (either abstractly or explicitly) between a set of taxa
that labels some of its nodes'' (although more narrow and context-dependent definitions can also
be found in the literature). Explicit rooted phylogenetic networks are rooted DAGs,
with leaves labelled by the taxa (species, groups, individuals or sequences), and nodes classed
as either reticulation nodes (with in-degree $\geq 2$) or tree nodes.

@@

% it's not reasonable to assume that one type of event happens
% at a time any more.
The eARG is also limited in the types of ancestral history that can
be described. This ARG encoding is derived from the coalescent
with recombination, which explicitly assumes that only one type
of event can occur (i.e., common ancestry or recombination) at a
time. This model is derived as an approximation to a Wright-Fisher
model, assuming that the sample size $n$ is much less than the
population size $N$, and that the genome is short enough
that the number of extant lineages remains
much smaller than $N$ at all times. These approximations are
reasonable to make for small datasets, such as
the classic \citet{kreitman1983nucleotide} \emph{Drosophila melanogaster}
ADH gene dataset (comprising 11 genomes over 2.7 kb of sequence with 43 SNPs,
see the ``\nameref{ARG_inference_methods}'' section). However,
they are highly questionable in modern datasets. % absurd really
For example, many
human datasets consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and thus our sample size is often much \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
% TODO make this better. Ask Gregor for some refs?
% Basically, Ag datasets make the assumption of isolated
% rare recombinations absurd.
Agricultural datasets such as the 1000 bull genomes~\citep{hayes20191000}
combine genome sequence data with detailed pedigrees, with
sample data from multiple generations.
Complete chromosome-level assemblies are now possible
in humans~\citep{miga2020telomere},
and projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
While the coalescent model can be suprisingly robust to such
basic violations to the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.
In a Wright-Fisher model, coalescence and recombination can
occur at the same time in the same genomes, and we can be
certain that this is also true in the incredibly rich
datasets that are available today.

A final limitation of the eARG encoding illustrated in
Fig.~\ref{fig-arg-data-structure}A is the inability to represent
multiple recombinations along a chromosome, gene conversion,
or transmission of multiple chromosomes.
[Sentence saying ``gene conversion is important, actually, cite, cite,
so we should be hoping to infer it some day''. Presumably people
already are in bacterial contexts?]
To allow us to encode data from these important processes we
must either devise workarounds to the event based encoding
(i.e., model gene conversion as two recombination events) or
generalise the model to more appropriately reflect the realities
of modern datasets.

@@

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
The Ancestral Recombination Graph defined in terms of genomes embedded
in a pedigree. (A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $I_1$ to $I_8$, contain both paternal (left) and maternal (right) genomes
labelled \textsf{a} to \textsf{p}. Black lines show transmission paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Two recombination breakpoints, at positions 2 and 7 mean that genomes \textsf{a}
and \textsf{c} are independent recombinants of the paternal genomes \textsf{e}
and \textsf{f}: in these cases the edge annotations have been shown, indicating the
region of genome transmitted. Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker; this shows that all
genomic regions are fully coalesced in genome \textsf{n} in the oldest generation.
(B) The equivalent genome ARG, where there is a single type of node (representing
a genome) along with annotations on all edges; annotations showing partial
transmission are highlighted in bold. Recombination, such as that causing the
breakpoint at position 7, is associated with two parent genomes (\textsf{e} and
\textsf{f}) and a child genome (the recombinant, \textsf{c}).
(C) The edge-annotated genome ARG in (B) fully defines the
local relationships along the genome, as shown in the local
or ``marginal'' trees.
}
\end{figure}

To distinguish from the classical event-based encodings,
we call this encoding of an ARG data
structure a ``Genome ARG'', or gARG.
Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG.
Fig.~\ref{fig-arg-in-pedigree}A shows the genomes
embedded in an inbred diploid pedigree, in which individuals occur at
discrete generations and carry two genomes.
% Each individual (labelled
% $I_1$ to $I_8$) therefore has a maternal and paternal genome,
% although the pedigree is only shown for concreteness, and
% there is no inherent limitation on mating system, ploidy, or age structure.
A genome is inherited from one of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{e} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination. Fig.~\ref{fig-arg-in-pedigree}B shows the resulting gARG,
depicting the ancestry of the sample genomes \textsf{a}--\textsf{d}. The graph
topology along with the genome coordinate annotations on the
edges then uniquely defines the trees at every position
along the genome (Fig.~\ref{fig-arg-in-pedigree}C).

@@

The focus on genomes rather than events has some interesting
and important ramifications. Firstly, a gARG may contain nodes that
do not correspond to any particular event in the ancestry of the samples.
For example, node \textsf{h} is the direct ancestor of \textsf{d} in
Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
coalescence or a recombination. Such nodes would usually be removed
from the graph so that \textsf{d} descends directly from \textsf{k} (see
the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
a representational perspective. This ability can be useful in practice and could be
% Do we need to spell this out?
required if, for example, we have a deeply sequenced pedigree of haplodiploids.
% e.g. Honeybees? Be great to cite an example!
% Note by Yan - if we actually have the sequence, then this turns into
% a sample node, so would be normally included anyway. Perhaps, therefore
% we should talk about these "pass though" nodes in the context
% of internal samples.

@@

The final point deals with the way in which recombination is represented, and is
closely linked to which genomes are chosen as the nodes in a gARG.
We could potentially choose genomes at any point in the life cycle: for instance,
we could chose to represent gametes as ARG nodes. However, the most intuitive
choice is where genomes represent diploid individuals in each generation, as in
Fig~\ref{fig-arg-in-pedigree}A. In this case recombination
is not represented by a single ``recombination node'' (as in an eARG), but with
two parent genomes between which recombination takes place (e.g.
nodes \textsf{e} and \textsf{f}), and a
single resulting ``recombinant'' genome (e.g. \textsf{c}). The
edges that link this recombinant with its two parents carry the information
concerning breakpoint location that would normally be present on the
recombination node in an eARG. For example, node \textsf{c} in
Fig~\ref{fig-arg-in-pedigree} inherits genome positions 0 to 7 from \textsf{e}
and positions 7 to 10 from \textsf{f}, indicating that the breakpoint is at position 7.

@@

The definition of an ARG as a data structure in terms of
individuals' monoploid genomes
and their direct patterns of inheritance allows us to describe arbitrarily
complex relatedness structures, free from the limitations imposed by
coalescent approximations.
An ARG is an intrinsically retrospective
structure, defined with respect to the ancestry of
a set of sampled genomes.
The graph structure (the nodes and their connectivity through edges)
captures a great deal of information about this ancestry,
but on its own it cannot tell us about the precise relationships between
the samples along the genome: do to this, we must construct the
local (or ``marginal'') trees.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
The \citet[][fig. 1]{wiuf1999recombination} example graph as a gARG, with contrasting
edge annotation schemes.
(A) Edge annotations that reflect immediate inheritance, i.e. transmission of all
genetic material between parent and child. All edges span the full genome from position 0 to 7,
apart from those linking a recombinant and its two parents (bold).
(B) Sample-resolved edge annotations: the Hudson simplification algorithm has been used to
shorten edge spans to reflect only the upwards transmission of genetic
material from the samples \textsf{a}, \textsf{b}, and \textsf{c}. The depiction of the genome within
each node is shaded as in figure \ref{fig-arg-in-pedigree}a to show
coalescing regions of genome back to the grand MRCA genome \textsf{w}.
In this case, using line widths to reflect the genomic span of each edge can
help to visualise the relative importance of different transmission pathways.
}
\end{figure}

@@

The state-space of the Griffiths process is much simpler than Hudson's algorithm,
which greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestry. The big ARG is vast, and any realisation
for even moderate levels of recombination is far too large to be of practical
use. The number of events in the big ARG, all the way back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the little ARG is
$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
This disparity in the number of events in the two formulations is
because the majority of the events that occur in the big ARG do
not affect the genetic ancestry of the sample in any way. Recombination
events that occur outside of ancestral material do not have any bearing
on the ancestry of the sample, and so the structure is hugely redundant.
As~\cite{wiuf1999recombination} note,
``an `ancestral' sequence in the birth and death process
need not have any genetic material in common with a
sequence descended from it.''
% Full quote:
% This process simplifies mathematics on the account that the notion of an
% ancestor will have a less restrictive meaning than usual:
% An ``ancestral'' sequence in the birth and death process
% need not have any genetic material in common with a
% sequence descended from it.

@@

Not all nodes in an ARG are equal. Assuming that we have sequenced the
genomes of the samples, there are intrinsic limits on what can
be known about more ancient nodes, determined by the graph topology
and the flow ancestral material through it.
\emph{Simplification}~\citep{kelleher2018efficient} is the process
of removing nodes and re-writing edges in a gARG to remove
various types of redundancy, while maintaining the crucial properties
that the meaningful topology in the local trees is not affected,
and the relationship between nodes in adjacent trees is maintained.

@@

The ideas of ``meaningful'' local tree topologies revolve around the
presence of ``unary'' nodes: a node in the tree that has exactly
one child. Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a Wright-Fisher
model, shown both as a graph on the left, and the equivalent set of 6
local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
\textsf{a} in the penultimate tree, highlighted in yellow and covering genomic positions
60 to 70. On this path, before $A$ meets its common ancestor with the other samples at
node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and
\textsf{p}, which in this region of the genome are all unary nodes. This reflects the
passage of this span of genome through the ARG towards the root (see the
``\nameref{Ancestry_resolution}'' section).
Such unary nodes in a local tree are not usually considered important,
because they provide very little information and are not
detectable by information from the tree. % weak
The nodes in a tree usually mark the points where samples find a
 \emph{common} ancestor, such as node \textsf{q} in the highlighted
tree. For many purposes, we would therefore consider a path directly
joining sample \textsf{a} to node \textsf{q} to be of equivalent status to the path
containing the intermediate unary nodes. Simplification is the process of removing
these less informative and (for most computational purposes) inefficient nodes
from the local trees and the corresponding annotated graph.

@@

\section*{ARGs and SPRs}
\citet{wiuf1999ancestry,wiuf1999recombination} recast
the coalescent with recombination
as a stochastic process along the genome rather than backwards in time.
In this formulation we begin at one end of a chromosome by
sampling a tree under the standard coalescent, and then reason
about how this tree is modified by recombination events
as we move along the genome. Although this algorithm is
significantly more complicated to simulate than Hudson's
backwards in time approach, it has led to substantial
insights into the process of recombination itself
and to the sequentially Markov Coalescent (SMC)
approximations~\citep{mcvean2005approximating,marjoram2006fast}
which are central to many modern methods.

@@

The conditional sampling probability $p(G | D)$ of an ARG $G$ given an observed
realization of genetic diversity $D$ at its sampled leaves is a central concept in
model fitting and selection. Under the Bayesian paradigm, it defines the posterior
distribution $p(G | D) \propto p(D | G) p(G)$, where $p(D | G)$ is the likelihood
of the data $D$ given $G$, and $p(G)$ is a prior distribution for the underlying
ARG $G$. Under the frequentist paradigm, $p(G)$ can be interpreted as a regularizer,
and $p(G | D)$ as a regularized likelihood. In both cases, $p(G)$ is typically specified
as the sampling distribution of a stochastic process, such as the coalescent with
recombination. Essentially all of the model-based inference methods surveyed below
rely on evaluation of a likelihood and a prior. Two contemporary examples of
likelihood-based methods are  \cite{mahmoudi2022bayesian, guo2022recombination}.

In essentially all neutral cases in practical use, the likelihood $p(D | G)$ is the law of
a Poisson process on the ancestral material on the edges of the ARG, which implies
that mutation processes at two sites are conditionally independent given the
underlying ancestral trees at those sites. A prototypical example is
\citet[Eq.\ (2)]{mahmoudi2022bayesian}. Hence, $p(D | G)$ can be evaluated
regardless of the precise data structure used to store $G$, provided that it encodes
at least all marginal ancestral trees along the sampled sequences.  Choosing the
right data structure can still have a dramatic effect on the efficiency with which
$p(D | G)$ can be evaluated \citep{mahmoudi2022bayesian}.

>>>>

% NOTE: it might be helpful to have a drawing of the ARG process
% in-progress here to help us talk about it. What is the full state
% that is remembered as we go through the process? It does help to think
% about the process as choosing edges to merge, not nodes.
% We could also refer to Fig.~\ref{fig-arg-data-structure}B if we want a simple picture
% to go with this
In the Griffiths formulation, each edge in the graph corresponds to an extant
lineage and nodes are events in the process (the initial $n$ leaf nodes are
``sampling'' events). Common ancestor events occur at rate $\binom{k}{2}$ when there
are $k$ lineages present. We choose two lineages (edges) uniformly, and merge them
into a common ancestor lineage. Recombination events happen at a rate of
$k \rho$. Here, we choose a lineage (edge) uniformly, and a
breakpoint $0 < x < m$ uniformly on its genome. We terminate the edge at a
node, record the breakpoint and start two new edges from this node. The process
then continues until there is only one lineage left (the Grand Most Recent
Common Ancestor, GMRCA), which is guaranteed to
happen in finite time because of the linear vs quadratic rates of branching
and coalescing. Later, we will show how the graph structure and breakpoints
associated with
recombination nodes provides sufficient information to later recover the marginal
trees and calculate sampling probabilities. However, this process traces out
a larger graph (the ``big'' ARG) than the Hudson algorithm (the ``little'' ARG; see the ``\nameref{Ancestry_resolution}'' section),
and is consequently considerably less efficient \citep{wiuf1999ancestry}.

@@

The idea of using branching-coalescing stochastic processes
to model genetic ancestral trees has also found application in natural selection.
The Ancestral Selection Graph (ASG)~\citep{krone1997ancestral,neuhauser1997genealogy}
uses dynamics identical to the ``big'' ARG to simulate an ensemble of correlated
potential ancestral trees. Weak genic selection is incorporated by sampling a
true ancestry from the ensemble in a non-uniform way.
There is no ASG analogue of the more computationally tractable ``little'' ARG,
though some gains in tractability can be made by considering typed
lineages~\citep{etheridge2009coalescent} or by leveraging perfect simulation
techniques when recurrent mutation is present \citep{fearnhead2001perfect}.
Extensions of the ASG have been developed to frequency-dependent
selection~\citep{neuhauser1999ancestral, gonzalezcasanova2018duality},
unlinked chromosomes~\citep{fearnhead2003ancestral}, recombining
loci whereupon branching is due to both selection and
recombination~\citep{donnelly1999genealogical}, and high fecundity
reproduction~\citep{gonzalezcasanova2018duality, koskela2019robust}.

@@

The term \emph{phylogenetic network} has been broadly accepted to describe the most general
type of genealogical graph, defined by \citet{huson2010phylogenetic} as ``any graph used to
represent evolutionary relationships (either abstractly or explicitly) between a set of taxa
that labels some of its nodes'' (although more narrow and context-dependent definitions can also
be found in the literature). Explicit rooted phylogenetic networks are rooted Directed Acyclic Graphs (DAGs),
with leaves labelled by the taxa (species, groups, individuals or sequences), and nodes classed
as either reticulation nodes (with in-degree $\geq 2$) or tree nodes.

@@

% it's not reasonable to assume that one type of event happens
% at a time any more.
The eARG is also limited in the types of ancestral history that can
be described. This ARG encoding is derived from the coalescent
with recombination, which explicitly assumes that only one type
of event can occur (i.e., common ancestry or recombination) at a
time. This model is derived as an approximation to a Wright-Fisher
model, assuming that the sample size $n$ is much less than the
population size $N$, and that the genome is short enough
that the number of extant lineages remains
much smaller than $N$ at all times. These approximations are
reasonable to make for small datasets, such as
the classic \citet{kreitman1983nucleotide} \emph{Drosophila melanogaster}
ADH gene dataset (comprising 11 genomes over 2.7 kb of sequence with 43 SNPs,
see the ``\nameref{ARG_inference_methods}'' section). However,
they are highly questionable in modern datasets. % absurd really
For example, many
human datasets consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and thus our sample size is often much \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
The 1000 bull genomes project~\citep{hayes20191000}
comprises close to 7000 genomes, which are part of multi-generation pedigrees
with millions of animals and extensive SNP array genotype data. For example,
the US dairy cattle database alone comprises more than 6 million animals with SNP array genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
while the effective population size in modern dairy cattle breeds is less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
This low effective population size is caused by intense selection and
staggering use of a handful of bulls across the world. The current record
holder is bull Toystory that has produced 2.4 million doses of semen
for artificial insemination across 13 years\footnote{\url{https://www.dairyherd.com/news/world-renowned-toystory-bull-has-died}}.
An extreme sign of these breeding practices is that there are only two ancestral
Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
Similar datasets are becoming available also in other livestock species.
Notably, \citet{RosFreixedes2022} have combined genome
and SNP array data to infer recombination of genomes for 440,610 pigs within
7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}.
Effective population size in pigs is also less than 100 with each female and male
routinely producing respectively more than 10 and 5000 progeny each \citep{Hall2016,Porcnic2016}.
Complete chromosome-level assemblies are now possible
in humans~\citep{miga2020telomere},
and projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
While the coalescent model can be suprisingly robust
basic violations of the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.
In a Wright-Fisher model, coalescence and recombination can
occur at the same time in the same genomes, which is increasingly
captured in the incredibly rich datasets that are available today.

A final limitation of the eARG encoding illustrated in
Fig.~\ref{fig-arg-data-structure}A is the inability to represent
multiple recombinations along a chromosome, gene conversion,
or transmission of multiple chromosomes.
[Sentence saying ``gene conversion is important, actually, cite, cite,
so we should be hoping to infer it some day''. Presumably people
already are in bacterial contexts? Gregor: I recall this paper from
Arabidopsis https://www.pnas.org/doi/10.1073/pnas.1211827110]
To allow us to encode data from these important processes we
must either devise workarounds to the event based encoding
(i.e., model gene conversion as two recombination events) or
generalise the model to more appropriately reflect the realities
of modern datasets.

@@

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
The Ancestral Recombination Graph defined in terms of genomes embedded
in a pedigree. (A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $I_1$ to $I_8$, contain both paternal (left) and maternal (right) genomes
labelled \textsf{a} to \textsf{p}. Black lines show transmission paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Two recombination breakpoints, at positions 2 and 7 mean that genomes \textsf{a}
and \textsf{c} are independent recombinants of the paternal genomes \textsf{e}
% Gregor: there is some overlap between a and c still (the middle region inherited
% from f) so, a and c are not fully independent!
and \textsf{f}: in these cases the edge annotations have been shown, indicating the
region of genome transmitted. Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker; this shows that all
genomic regions fully coalesce in genome \textsf{n} in the oldest generation.
(B) The equivalent genome ARG, where there is a single type of node (representing
a genome) along with annotations on all edges; annotations showing partial
transmission are highlighted in bold. Recombination, such as that causing the
breakpoint at position 7, is associated with two parent genomes (\textsf{e} and
\textsf{f}) and a child genome (the recombinant, \textsf{c}).
(C) The edge-annotated genome ARG in (B) fully defines the
local relationships along the genome, as shown in the local
(``marginal'') trees.
}
\end{figure}

To distinguish from the classical event-based encodings,
we call this encoding of an ARG data
structure a ``Genome ARG'', or gARG.
Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG.
Fig.~\ref{fig-arg-in-pedigree}A shows the genomes
embedded in an inbred diploid pedigree, in which individuals occur at
discrete generations and carry two genomes. Each individual (labelled
$I_1$ to $I_8$) therefore has a paternal genome (on the left) and a maternal one
(on the right), although the pedigree is only shown for concreteness, and
there is no inherent limitation on mating system, ploidy, or age structure.
A genome is inherited from one of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination. Fig.~\ref{fig-arg-in-pedigree}B shows the resulting gARG,
depicting the ancestry of the sample genomes \textsf{a}--\textsf{d}. The graph
topology along with the genome coordinate annotations on the
edges then uniquely defines the trees at every position
along the genome (Fig.~\ref{fig-arg-in-pedigree}C).

@@

The focus on genomes rather than events has some interesting
and important ramifications. Firstly, a gARG may contain nodes that
do not correspond to any particular event in the ancestry of the samples.
For example, node \textsf{h} is the direct ancestor of \textsf{d} in
Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
coalescence or a recombination. Such nodes would usually be removed
from the graph so that \textsf{d} descends directly from \textsf{k} (see
the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
a representational perspective. This ability can be useful in practice and could be
% Do we need to spell this out? Gregor: I think so - see below
required if, for example, we have a deeply sequenced pedigree as in \cite{RosFreixedes2022}.
% e.g. Honeybees? Be great to cite an example!
% Note by Yan - if we actually have the sequence, then this turns into
% a sample node, so would be normally included anyway. Perhaps, therefore
% we should talk about these "pass though" nodes in the context
% of internal samples.
% Gregor: There are no good deeply sequenced pedigrees in honeybees, but sex
% chromosome behaves like haplodiploid system, also any deep pedigree works here, so I cited our pig
% work. Also, I think its important to not think of all genomes as current samples - we have many
% generations of animals with genome or SNP array data and we know pedigree so we can't assume they
% are all current samples - I believe you made this clear distinction also in the tsdate paper with
% ancient human samples

@@

The final point deals with the way in which recombination is represented, and is
closely linked to which genomes are chosen as the nodes in a gARG.
We could potentially choose genomes at any point in the life cycle: for instance,
we could chose to represent gametes as ARG nodes. However, the most intuitive
choice is where genomes represent diploid individuals in each generation, as in
Fig~\ref{fig-arg-in-pedigree}A. In this case recombination
is not represented by a single ``recombination node'' (as in an eARG), but with
two parent genomes between which recombination takes place (e.g.
nodes \textsf{e} and \textsf{f}), and a
single resulting ``recombinant'' genome (e.g. \textsf{c}). The
edges that link this recombinant with its two parents carry the information
concerning breakpoint location that would normally be present on the
recombination node in an eARG. For example, node \textsf{c} in
Fig~\ref{fig-arg-in-pedigree} inherits genome positions 0 to 7 from \textsf{f}
and positions 7 to 10 from \textsf{e}, indicating that the breakpoint is at position 7.

@@

The definition of an ARG as a data structure in terms of
individuals' monoploid genomes
and their direct patterns of inheritance allows us to describe arbitrarily
complex relatedness structures, free from the limitations imposed by
coalescent approximations.
An ARG is an intrinsically retrospective
structure, defined with respect to the ancestry of
a set of sampled genomes.
The graph structure (the nodes and their connectivity through edges)
captures a great deal of information about this ancestry,
but on its own it cannot tell us about the precise relationships between
the samples along the genome: do to this, we must construct the
local (``marginal'') trees.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
The \citet[][fig. 1]{wiuf1999recombination} example graph as a gARG, with contrasting
edge annotation schemes.
(A) Edge annotations that reflect immediate inheritance, i.e. transmission of
genetic material between parent and child. All edges span the full genome from position 0 to 7,
apart from those linking a recombinant and its two parents (bold).
(B) Sample-resolved edge annotations: the Hudson simplification algorithm has been used to
shorten edge spans to reflect only the upwards transmission of genetic
material from the samples \textsf{a}, \textsf{b}, and \textsf{c}. The depiction of the genome within
each node is shaded as in figure \ref{fig-arg-in-pedigree}a to show
coalescing regions of genome back to the grand MRCA genome \textsf{w}.
In this case, using line widths to reflect the genomic span of each edge can
help to visualise the relative importance of different transmission pathways.
}
\end{figure}

@@

The state-space of the Griffiths process is much simpler than Hudson's algorithm,
which greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestry. The big ARG is vast, and any realisation
for even moderate levels of recombination is far too large to be of practical
use. The number of events in the big ARG all the way back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the little ARG is
$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
This disparity in the number of events in the two formulations is
because the majority of the events that occur in the big ARG do
not affect the genetic ancestry of the sample in any way. Recombination
events that occur outside of ancestral material do not have any bearing
on the ancestry of the sample, and so the structure is hugely redundant.
As~\cite{wiuf1999recombination} note,
``an `ancestral' sequence in the birth and death process
need not have any genetic material in common with a
sequence descended from it.''
% Full quote:
% This process simplifies mathematics on the account that the notion of an
% ancestor will have a less restrictive meaning than usual:
% An ``ancestral'' sequence in the birth and death process
% need not have any genetic material in common with a
% sequence descended from it.

@@

Not all nodes in an ARG are equal. Assuming that we have sequenced the
genomes of the samples, there are intrinsic limits on what can
be known about more ancient nodes, determined by the graph topology
and the flow of ancestral material through it.
\emph{Simplification}~\citep{kelleher2018efficient} is the process
of removing nodes and re-writing edges in a gARG to remove
various types of redundancy, while maintaining the crucial properties
that the ``meaningful'' topology in the local trees is not affected,
and the relationship between nodes in adjacent trees is maintained.

@@

The ideas of ``meaningful'' local tree topologies revolve around the
presence of ``unary'' nodes: a node in the tree that has exactly
one child. Fig.~\ref{fig-simplification} show a series of successive steps of
simplification, starting with a complete gARG simulated under a Wright-Fisher
model, shown both as a graph on the left, and the equivalent set of 6
local trees on the right (Fig.~\ref{fig-simplification}A). Consider the path above node
\textsf{a} in the penultimate tree, highlighted in yellow and covering genomic positions
60 to 70. On this path, before \textsf{a} meets its common ancestor with the other samples at
node \textsf{q}, it passes through  \textsf{e},  \textsf{j},  \textsf{m},  \textsf{n} and
\textsf{p}, which in this region of the genome are all unary nodes. This reflects the
passage of this span of genome through the ARG towards the root (see the
``\nameref{Ancestry_resolution}'' section).
Such unary nodes in a local tree are not usually considered important,
because they provide very little information and are not
detectable by information from the tree. % weak
The nodes in a tree usually mark the points where samples find a
 \emph{common} ancestor, such as node \textsf{q} in the highlighted
tree. For many purposes, we would therefore consider a path directly
joining sample \textsf{a} to node \textsf{q} to be of equivalent status to the path
containing the intermediate unary nodes. Simplification is the process of removing
these less informative and (for most computational purposes) inefficient nodes
from the local trees and the corresponding annotated graph.

@@

\section*{ARGs and SPRs}
\citet{wiuf1999ancestry,wiuf1999recombination} recast
the coalescent with recombination
as a stochastic process along the genome rather than backwards in time.
In this formulation we begin at one end of a chromosome by
sampling a tree under the standard coalescent, and then reason
about how this tree is modified by recombination events
as we move along the genome. Although this algorithm is
significantly more complicated to simulate than Hudson's
backwards in time approach, it has led to substantial
insights into the process of recombination itself
and to the sequentially Markov Coalescent (SMC)
approximations~\citep{mcvean2005approximating,marjoram2006fast}
that are central to many modern methods.

@@

The conditional sampling probability $p(G | D)$ of an ARG $G$ given an observed
realization of genetic diversity $D$ at its sampled leaves is a central concept in
model fitting and selection. Under the Bayesian paradigm, it defines the posterior
distribution $p(G | D) \propto p(D | G) p(G)$, where $p(D | G)$ is the likelihood
of the data $D$ given $G$, and $p(G)$ is a prior distribution for the underlying
ARG $G$. Under the frequentist paradigm, $p(G)$ can be interpreted as a regularizer,
and $p(G | D)$ as a regularized likelihood. In both cases, $p(G)$ is typically specified
as the sampling distribution of a stochastic process, such as the coalescent with
recombination. Essentially all of the model-based inference methods surveyed below
rely on evaluation of a likelihood and a prior. Two contemporary examples of
likelihood-based methods are \cite{mahmoudi2022bayesian} and \cite{guo2022recombination}.

In essentially all neutral cases in practical use, the likelihood $p(D | G)$ is based on
a Poisson process on the ancestral material on the edges of the ARG, which implies
that mutation processes at two sites are conditionally independent given the
underlying ancestral trees at those sites. A prototypical example is
\citet[Eq.\ (2)]{mahmoudi2022bayesian}. Hence, $p(D | G)$ can be evaluated
regardless of the precise data structure used to store $G$, provided that it encodes
at least all marginal ancestral trees along the sampled sequences.  Choosing the
right data structure can still have a dramatic effect on the efficiency with which
$p(D | G)$ can be evaluated \citep{mahmoudi2022bayesian}.

In contrast, evaluating $p(G)$ under the coalescent with recombination
(or related models) requires knowledge of the number of extant lineages and
the number of links available for recombination (i.e.\ ones at which a recombination
would split ancestral material) at each time along the ARG
\citep[Eq.\ (3)]{mahmoudi2022bayesian}. These cannot be uniquely determined
from just the marginal trees at each site because branches at multiple trees can be
contained in the same ancestral lineage. Instead, a richer data structure akin to the
eARG or gARG is needed. Moreover, because eARG does not explicitly store the flow
of ancestral material along its edges---a necessary piece of information for
computing the number of effective recombination links at each time---evaluating
$p(G)$ based on an eARG essentially requires turning it into a gARG at runtime.
